FITS format for planetary surfaces: definitions, applications and best practices.

### 2018-08-22 ess_marmo.tex
Adds acronym table

In this paper we will describe how a small addition to the Flexible Image Transport
System (FITS) standard, widely used in the astronomy investigations, will allow FITS
to be more easily used in planetary surface investigations.
We will also show how FITS metadata can easily be transformed
in the PDS4 metadata archival model and lastly provide example implementations.
More than imposing a formal data model, a FITS description for planetary data aims
to simplify sharing data across the planetary and astronomy domains and promoting
interoperability from raw data formatting to final visualization.
\end{abstract}

@@

This paper proposes to use Flexible Image Transport
System\footnote{https://fits.gsfc.nasa.gov/}
(FITS) format \citep{fitsorig,fitsver3} in planetary surface investigations,
and describes how FITS metadata can easily be inserted in the
PDS4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml}
metadata distribution model.
FITS is one of the standard formats implemented in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO), therefore making obvious its connection to the Planetary Virtual
Observatory initiative and the planetary Table Access Protocol (EPN-TAP)
\citep{ERARDepntap}.
FITS format is open, flexible, and largely implemented in open and efficient
processing tools allowing to handle large amounts of raster data.
The goal of this approach is to make easier data mining and reprocessing
in planetary surface investigations, promoting general software based on those
standards.
The option to use FITS within the planetary surface domain can help to
homogenize methods from raw formatting to visualization, while
allowing for optimized data processing across the planetary and astronomy domains.

@@

\subsection{FITS Format}
FITS is an open digital standard, created in the late 1970's for data acquisition,
transfer and archiving of telescope data by astronomical observatories, where
it has been used during the last thirty years.
It had been adopted for data exchange and archiving from several orbital telescopes
and spatial missions.
The International Astronomical Union approved FITS as the standard format
for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard formats implemented in the Virtual
Observatory (VO).
FITS data storage is compatible with the Planetary Data System (PDS)
archiving specifications so that FITS files can be embedded in PDS datasets.
Finally, it is supported by a large number of open libraries
and software tools.

FITS is already capable of being used as a standard formatting for most data products
commonly used in planetary surface investigations.
In particular, the Multi-Extension FITS (MEF) schema proposes an easy way
to store inhomogeneous digital information (reflectance, calibration data,
vector, table data, etc.) in the same file, each with corresponding metadata,
as well as multi-detector imagery (e.g. from HiRISE \citep{hirise})
or hyperspectral cubes with geometry information (e.g. from CRISM
\citep{crism} or OMEGA \citep{omega} instruments).
FITS has been already chosen to distribute data from, e.g.,
Hayabusa AMICA and NIRS cameras\footnote{http://darts.isas.jaxa.jp/planet/project/hayabusa/index.html},
all Akatsuki cameras\footnote{http://darts.isas.jaxa.jp/planet/project/akatsuki/index.html}
(except the Lightning and Airglow Camera),
and the Dawn Framing Camera\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
Some Rosetta data, from NAVCAM and OSIRIS cameras, are distributed in FITS format at
the ESA Planetary Science Archive\footnote{http://psa.esa.int} (PSA).
In addition, many PDS3 datasets in the Small Bodies Node are archived and distributed
as FITS files with PDS3 labels.
However, to be more efficiently used in planetary surface investigations, FITS metadata
must be extended in order to take into account the size of the reference body and
orientation as standardized by the IAU Working Group on Cartographic Coordinates and
Rotational Elements (WGCCRE).
This group reports triennially \citep{wgccre} on the preferred rotation rate,
spin axis, prime meridian, and reference surface for planets and satellites
ensuring that cartographic endeavors are effectively comparable.
In the framework of the VESPA\footnote{http://europlanet-vespa.eu/}
\citep{vespapss} workpackage of the EuroplanetRI2020\footnote{http://www.europlanet-2020-ri.eu/}
project an extension to FITS metadata (GeoFITS) has been
proposed\footnote{https://voparis-confluence.obspm.fr/display/VES/GeoFITS:+Planetary+Data+FITS+format+and+metadata+convention},
taking into account the WGCCRE recommendations.

@@

FITS World Coordinate System
representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html} is the standard
way to describe spatial dependencies in FITS metadata.
WCS simplifies the spatial coordinate description with respect to historical
terrestrial references.
For instance, there is no need for oblique projection definitions
\citep{calagreisenI,Snyder1987} as projection parameters
(projection center or reference point)
describe the difference between, e.g., the Mercator and
Oblique Mercator or Simple Cylindrical and
Equirectangular map projections.
This simplification is effective when a spherical local approximation
is applied on planetary imagery.
This is particularly the case with growing image resolutions and
increasing computing capabilities: spherical projections with a local
radius definition are easier to reproject from one system to another, than
from or to a global spheroidal datum.
As an example, the HiRISE team has chosen to distribute HiRISE projected
products using Equirectangular projection (CAR) on a local surface, approximated
by the sphere with radius equivalent to the local Mars spheroid radius.

@@

\subsubsection{FITS WCS keywords and their recommended use in planetary surface context}
\label{sssec:wcskeys}
In this section WCS FITS keywords are inserted in the planetary surface context.
When possible, corresponding PDS4 and EPN-TAP definitions are added for archiving
and distribution standardization purposes.
\begin{enumerate}
\item{RADESYS.
Name of the inertial reference frame with respect to which the coordinate
system is defined. Standard FITS defaults this keyword to \texttt{`ICRS'}.
This is applicable to planetary data too as long as the planetary coordinate system
parameters, i.e. the rotational elements, are defined with respect to \texttt{`ICRS'}.
This is the case starting from the 2003 WGCCRE report \citep{wgccre2003}.
Before the 1982 WGCCRE report \citep{wgccre1982} the inertial system was \texttt{`FK4'},
after it was \texttt{`FK5'}.}
\item{EQUINOX.
The value field contains the equinox in years for the celestial
coordinate system in which positions are expressed.
Standard FITS defaults this keyword to \texttt{`J2000'}.
This is applicable to planetary data starting from the 1982 WGCCRE report \citep{wgccre1982}.
Before 1982 the standard equinox was \texttt{`B1950'}.}
\item{WCSAXES.
Number of axes in the WCS description.
The value of WCSAXES may exceed the number of raster pixel axes. WCSAXES is, e.g.,
generally set to $2$ in two-dimensional rasters, it could be set to $3$ if an
additional time or filter dependency is added to the coordinate description.}
\item{WCSNAME.
Name of world coordinate descriptions (e.g. body-fixed rotating -ocentric -ographic,
non-rotating, inertial).
EPN-TAP provides similar information through the \texttt{spatial\_coordinate\_description} parameter.}
\item{CRPIXn.
Location of a reference point along axis $n$, in units of the axis index.
The reference point value do not need to lie within the actual data array.}
\item{CTYPEn.
Name of the coordinate represented by axis $n$, in ``4--3'' form: the first four
characters specify the coordinate type, the fifth character is a `-' and the
remaining three characters specify an algorithm code for computing the world coordinate
value (\ref{table:proj}).
For angular coordinates \citet{calagreisenI} provides the form xLN xLT or yzLN yzLT,
where x or yz correspond to the target body.
We propose to standardize those codes as listed in Table \ref{table:ctype}.
For projected coordinates in meters we propose a similar scheme: xPX as Projected X,
xPY as Projected Y, or yzPX and yzPY.
Alternative representation of WCS are already possible in FITS, to describe images in
degrees and in meters \citep{greisencalaII}.
The name of the alternative axis will be specified using the alternative CNAMEn
keyword.}
\item{CRVALn.
Value of the coordinate specified by the CTYPEn keyword at the reference point CRPIXn.}
\item{CDi\_j.
The $ij$ linear transformation matrix (rotation, skewness, scaling) defining the
intermediate world coordinates.}
\item{PCi\_j.
The $ij$ linear transformation matrix (rotation, skewness) defining the intermediate
pixel coordinates (to be used with CDELT keywords).}
\item{CDELTn.
The coordinate increment along axis $n$ (scaling component of the linear transformation
matrix PCi\_j).
They correspond to the \textit{pixel\_resolution\_x} and
\texttt{pixel\_resolution\_y} attributes in PDS4, to the \texttt{spatial\_resolution} parameter
in EPN-TAP.}
\item{CUNITn.
Units of the coordinate system. This is a not mandatory keyword: when not present FITS standard
assumes its value as degrees, the most common description in astronomical sky images.
The same can be assumed for planetary surfaces: when linear coordinates are used SI units must
be used (meters, m).}
\item{CNAMEn.
The description of a particular coordinate for an axis in a particular WCS (up to 68
characters), while the WCSNAMEa keyword names the particular WCS as a whole.
Its default value will be blank.}
\end{enumerate}

@@

\begin{sidewaystable}
\caption{Proposed keywords are related to 3D surface shape,
not available in astronomical celestial standards.}
\centering
\begin{tabular}{l l l l}
\hline
Keyword & Type & Status & Definition  \\
\hline
WGCCRECS & string & reserved & Value field contains a string referring to the WGCCRE report \\
& & & or document defining the coordinate system describing the data. (DOI)    \\
A\_RADIUS$^{a}$ & real & mandatory & Value field contains the semi-major axis of the ellipsoid that \\
& & & defines the approximate shape of a target body used in projection. \\
& & & `A' is usually in the equatorial plane. Always in meters.  \\
B\_RADIUS$^{a}$ & real & mandatory & Value field contains the value of the intermediate axis of \\
& & & the ellipsoid that defines the approximate shape of a target body used in projection. \\
& & & `B' is usually in the equatorial plane. Always in meters.   \\
C\_RADIUS$^{a}$ & real & mandatory & Value field contains the value of the semi-minor axis of the \\
& & & ellipsoid that defines the approximate shape of a target body used in projection. \\
& & & `C' is normal to the plane defined by `A' and `B'. Always in meters. \\
OGCCODE	& string & reserved, optional & Value field contains a string describing the standard OGC code \\
& & & (if any) from \citet{hare2006} corresponding to the shape and projection used in the image. \\
& & & This keyword will assure compatibility with standard GIS softwares. \\
\hline
\multicolumn{4}{l}{$^{a}$Note that IAU Working Group defines a mean radius for each Solar System body. When this Mean Radius is used in projection definition,} \\
\multicolumn{4}{l}{  the three keywords A\_RADIUS, B\_RADIUS,C\_RADIUS, must be set to the same value of the defined Mean Radius.} \\
\end{tabular}
\label{table:newkeys}
\end{sidewaystable}

@@

\item{\texttt{DATE-OBS}.
It contains the date of the observation, provided as ISO 8601
string. The value shall be assumed to refer to the start of the observation,
unless another interpretation is clearly explained in the comment field.
It is assumed to be expressed in UTC but other systems are accepted as long as they
are specified in the ISO string\footnote{See table 30 in the Standard FITS document https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf}. 
PDS4 provides a \texttt{Time\_Coordinates} class with attributes
\texttt{start\_date\_time} and \texttt{stop\_date\_time} providing start and
stop date and time appropriate to the product being labeled, always in UTC.
The \texttt{time\_min} and \texttt{time\_max} fields are mandatory in the EPN-TAP
scheme, they must always be formatted in UTC as a ISO string (planned in EPN-TAP v2.1).}

@@

\item{\texttt{REFERENC}.
It contains a character string citing the bibliographic reference where the
data are first described.
It is recommended that either the Astrophysics Data System bibliographic
databases\footnote{http://adswww.harvard.edu/}
bibliographic identifier or the Digital Object Identifier\footnote{http://doi.org}
be used.
PDS4 \texttt{External\_Reference} class and EPN-TAP \texttt{bib\_reference} have
exactly the same meaning and use than the FITS keyword.}

@@

\subsubsection{Visualization of Global and Projected Maps}
\label{sssec:maps}
High level products are meant to be used for visual inspection or GIS analysis.
They are distributed as projected rasters, often in
GeoTIFF\footnote{http://geotiff.osgeo.org/}
format, as GeoTIFF make them GIS ready.
Given some new keywords introduced in FITS and WCS definitions adapted to planetary
surface, FITS and GeoTIFF are completely interchangeable.
As an example we have downloaded a Mercury polar map from the USGS
PDS Annex\footnote{https://astrogeology.usgs.gov/search} and converted it in FITS.
The FITS image has been cropped in two smaller versions using the open source software
SWarp\footnote{https://www.astromatic.net/software/swarp} \citep{swarp}: SWarp was able to
automatically manage WCS planetary information.
We used ds9 to display the three FITS images and were able to correctly align them.
Linear coordinates in meters are visualized (see figure \ref{fig:ds9qgis}, left side)
as in QGIS, a popular open source GIS software, thanks to the alternate coordinate
description keywords in the header.
\begin{figure}[ht!]
\centerline{\includegraphics[height=17pc]{ess_marmo2} \includegraphics[height=17pc]{ess_marmo1}}
\caption{left: Mercury South Pole FITS images displayed in ds9, alternate coordinates in meters
are displayed;
right: GeoTIFF raster (displayed in red) and the two virtual haeders
linked to FITS rasters (displayed in green and blue); the three layers
are perfectly superposed.
Mercury image credits: MESSENGER Team, Arizona State University, Johns Hopkins Applied
Physics Laboratory, Carnegie Science}
\label{fig:ds9qgis}
\end{figure}

@@

\subsection{MATISSE}
MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool
for the Instruments for the Solar System Exploration) \citep{ZINZI2016} is the webtool
designed by the Space Science Data Center of the Italian Space Agency (SSDC-ASI) to
access, visualize and analyze data from solar system exploration missions.
Currently, MATISSE provides access to public data from five different targets (i.e., Mercury,
Venus, Mars, the dwarf planet Ceres and the asteroid Vesta).
MATISSE 2D outputs are formatted as FITS, in equirectangular projection and in planetocentric coordinates.
The planetary FITS standard described in this work has been implemented in MATISSE in order to make the output
file compliant with most FITS readers, especially DS9 and JS9.
The files generated by MATISSE could be also opened with standard GIS software (e.g.,
QGIS, ArcGIS) 
after being provided as input of the fits2vrt Python library\footnote{https://github.com/epn-vespa/fits2vrt},
whose output is a VRT virtual GDAL header pointing to the FITS file with the appropriate
projection.
Thanks to this approach, in the next major release of MATISSE (v2.0, planned to be
online by the end of 2018) JS9 will be the standard reader for
the 2D online output of the tool, thus allowing users to perform advanced analysis in
both 2D and 3D directly inside a web-based environment.

@@

We understand that some believe FITS has grown too old for newer astronomical
applications \citep{THOMAS2015133}. We are also not convinced that one format
can solve all the issues faced from long-term archiving to on-line streaming.
But we want to stress that using FITS can be an efficient compromise.
FITS can become the format that allows the astronomy and the  planetary science 
research communities to more easily share data across domains, allowing for
reduced data format conversions during processing and easy conversion for final
archival products.

@@

%
%%%%%%%%%%%%%%
% Acronyms
% \begin{acronyms}
% \acro{Acronym}
% Definition here
% \acro{EMOS}
% Ensemble model output statistics 
% \acro{ECMWF}
% Centre for Medium-Range Weather Forecasts
% \end{acronyms}

>>>>

In this paper we will describe how a small addition to the Flexible Image Transport
System (FITS) standard, widely used in the astronomy investigations, will allow FITS
to be more easily used in planetary surface investigations.
We will also show how FITS metadata can easily be transformed
in the Planetary Data System (PDS4) metadata archival model and lastly provide example 
implementations.
More than imposing a formal data model, a FITS description for planetary data aims
to simplify sharing data across the planetary and astronomy domains and promoting
interoperability from raw data formatting to final visualization.
\end{abstract}

@@

This paper proposes to use FITS\footnote{https://fits.gsfc.nasa.gov/}
format \citep{fitsorig,fitsver3} in planetary surface investigations,
and describes how FITS metadata can easily be inserted in the
PDS4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml}
metadata distribution model.
FITS is one of the standard formats implemented in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO), therefore making obvious its connection to the Planetary Virtual
Observatory initiative and the planetary Table Access Protocol (EPN-TAP)
\citep{ERARDepntap}.
FITS format is open, flexible, and largely implemented in open and efficient
processing tools allowing to handle large amounts of raster data.
The goal of this approach is to make easier data mining and reprocessing
in planetary surface investigations, promoting general software based on those
standards.
The option to use FITS within the planetary surface domain can help to
homogenize methods from raw formatting to visualization, while
allowing for optimized data processing across the planetary and astronomy domains.

@@

\subsection{FITS Format}
FITS is an open digital standard, created in the late 1970's for data acquisition,
transfer and archiving of telescope data by astronomical observatories, where
it has been used during the last thirty years.
It had been adopted for data exchange and archiving from several orbital telescopes
and spatial missions.
The International Astronomical Union (IAU) approved FITS as the standard format
for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard formats implemented in the Virtual
Observatory.
FITS data storage is compatible with the Planetary Data System (PDS)
archiving specifications so that FITS files can be embedded in PDS datasets.
Finally, it is supported by a large number of open libraries
and software tools.

FITS is already capable of being used as a standard formatting for most data products
commonly used in planetary surface investigations.
In particular, the Multi-Extension FITS (MEF) schema proposes an easy way
to store inhomogeneous digital information (reflectance, calibration data,
vector, table data, etc.) in the same file, each with corresponding metadata,
as well as multi-detector imagery (e.g. from HiRISE \citep{hirise})
or hyperspectral cubes with geometry information (e.g. from CRISM
\citep{crism} or OMEGA \citep{omega} instruments).
FITS has been already chosen to distribute data from, e.g.,
Hayabusa AMICA and NIRS cameras\footnote{http://darts.isas.jaxa.jp/planet/project/hayabusa/index.html},
all Akatsuki cameras\footnote{http://darts.isas.jaxa.jp/planet/project/akatsuki/index.html}
(except the Lightning and Airglow Camera),
and the Dawn Framing Camera\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
Some Rosetta data, from NAVCAM and OSIRIS cameras, are distributed in FITS format at
the ESA Planetary Science Archive\footnote{http://psa.esa.int} (PSA).
In addition, many PDS3 datasets in the Small Bodies Node are archived and distributed
as FITS files with PDS3 labels.
However, to be more efficiently used in planetary surface investigations, FITS metadata
must be extended in order to take into account the size of the reference body and
orientation as standardized by the IAU Working Group on Cartographic Coordinates and
Rotational Elements (WGCCRE).
This group reports triennially \citep{wgccre} on the preferred rotation rate,
spin axis, prime meridian, and reference surface for planets and satellites
ensuring that cartographic endeavors are effectively comparable.
In the framework of the Virtual European Solar and Planetary Access\footnote{http://europlanet-vespa.eu/} (VESPA)
\citep{vespapss} workpackage of the EuroplanetRI2020\footnote{http://www.europlanet-2020-ri.eu/}
project an extension to FITS metadata (GeoFITS) has been
proposed\footnote{https://voparis-confluence.obspm.fr/display/VES/GeoFITS:+Planetary+Data+FITS+format+and+metadata+convention},
taking into account the WGCCRE recommendations.

@@

FITS World Coordinate System (WCS)
representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html} is the standard
way to describe spatial dependencies in FITS metadata.
WCS simplifies the spatial coordinate description with respect to historical
terrestrial references.
For instance, there is no need for oblique projection definitions
\citep{calagreisenI,Snyder1987} as projection parameters
(projection center or reference point)
describe the difference between, e.g., the Mercator and
Oblique Mercator or Simple Cylindrical and
Equirectangular map projections.
This simplification is effective when a spherical local approximation
is applied on planetary imagery.
This is particularly the case with growing image resolutions and
increasing computing capabilities: spherical projections with a local
radius definition are easier to reproject from one system to another, than
from or to a global spheroidal datum.
As an example, the HiRISE team has chosen to distribute HiRISE projected
products using Equirectangular projection on a local surface, approximated
by the sphere with radius equivalent to the local Mars spheroid radius.

@@

\subsubsection{FITS WCS keywords and their recommended use in planetary surface context}
\label{sssec:wcskeys}
In this section WCS FITS keywords are inserted in the planetary surface context.
When possible, corresponding PDS4 and EPN-TAP definitions are added for archiving
and distribution standardization purposes.
\begin{enumerate}
\item{RADESYS.
Name of the inertial reference frame with respect to which the coordinate
system is defined. Standard FITS defaults this keyword to \texttt{`ICRS'}.
This is applicable to planetary data too as long as the planetary coordinate system
parameters, i.e. the rotational elements, are defined with respect to \texttt{`ICRS'}.
This is the case starting from the 2003 WGCCRE report \citep{wgccre2003}.
Before the 1982 WGCCRE report \citep{wgccre1982} the inertial system was \texttt{`FK4'},
after it was \texttt{`FK5'}.}
\item{EQUINOX.
The value field contains the equinox in years for the celestial
coordinate system in which positions are expressed.
Standard FITS defaults this keyword to \texttt{`J2000'}.
This is applicable to planetary data starting from the 1982 WGCCRE report \citep{wgccre1982}.
Before 1982 the standard equinox was \texttt{`B1950'}.}
\item{WCSAXES.
Number of axes in the WCS description.
The value of WCSAXES may exceed the number of raster pixel axes. WCSAXES is, e.g.,
generally set to $2$ in two-dimensional rasters, it could be set to $3$ if an
additional time or filter dependency is added to the coordinate description.}
\item{WCSNAME.
Name of world coordinate descriptions (e.g. body-fixed rotating -ocentric -ographic,
non-rotating, inertial).
EPN-TAP provides similar information through the \texttt{spatial\_coordinate\_description} parameter.}
\item{CRPIXn.
Location of a reference point along axis $n$, in units of the axis index.
The reference point value do not need to lie within the actual data array.}
\item{CTYPEn.
Name of the coordinate represented by axis $n$, in ``4--3'' form: the first four
characters specify the coordinate type, the fifth character is a `-' and the
remaining three characters specify an algorithm code for computing the world coordinate
value (see Table \ref{table:proj}).
For angular coordinates \citet{calagreisenI} provides the form xLN xLT or yzLN yzLT,
where x or yz correspond to the target body.
We propose to standardize those codes as listed in Table \ref{table:ctype}.
For projected coordinates in meters we propose a similar scheme: xPX as Projected X,
xPY as Projected Y, or yzPX and yzPY.
Alternative representation of WCS are already possible in FITS, to describe images in
degrees and in meters \citep{greisencalaII}.
The name of the alternative axis will be specified using the alternative CNAMEn
keyword.}
\item{CRVALn.
Value of the coordinate specified by the CTYPEn keyword at the reference point CRPIXn.}
\item{CDi\_j.
The $ij$ linear transformation matrix (rotation, skewness, scaling) defining the
intermediate world coordinates.}
\item{PCi\_j.
The $ij$ linear transformation matrix (rotation, skewness) defining the intermediate
pixel coordinates (to be used with CDELT keywords).}
\item{CDELTn.
The coordinate increment along axis $n$ (scaling component of the linear transformation
matrix PCi\_j).
They correspond to the \texttt{pixel\_resolution\_x} and
\texttt{pixel\_resolution\_y} attributes in PDS4, to the \texttt{spatial\_resolution} parameter
in EPN-TAP.}
\item{CUNITn.
Units of the coordinate system. This is a not mandatory keyword: when not present FITS standard
assumes its value as degrees, the most common description in astronomical sky images.
The same can be assumed for planetary surfaces: when linear coordinates are used SI units must
be used (meters, m).}
\item{CNAMEn.
The description of a particular coordinate for an axis in a particular WCS (up to 68
characters), while the WCSNAMEn keyword names the particular WCS as a whole.
Its default value will be blank.}
\end{enumerate}

@@

\begin{sidewaystable}
\caption{Proposed keywords are related to 3D surface shape,
not available in astronomical celestial standards.}
\centering
\begin{tabular}{l l l l}
\hline
Keyword & Type & Status & Definition  \\
\hline
WGCCRECS & string & reserved & Value field contains a string referring to the WGCCRE report \\
& & & or document defining the coordinate system describing the data. (DOI)    \\
A\_RADIUS$^{a}$ & real & mandatory & Value field contains the semi-major axis of the ellipsoid that \\
& & & defines the approximate shape of a target body used in projection. \\
& & & `A' is usually in the equatorial plane. Always in meters.  \\
B\_RADIUS$^{a}$ & real & mandatory & Value field contains the value of the intermediate axis of \\
& & & the ellipsoid that defines the approximate shape of a target body used in projection. \\
& & & `B' is usually in the equatorial plane. Always in meters.   \\
C\_RADIUS$^{a}$ & real & mandatory & Value field contains the value of the semi-minor axis of the \\
& & & ellipsoid that defines the approximate shape of a target body used in projection. \\
& & & `C' is normal to the plane defined by `A' and `B'. Always in meters. \\
OGCCODE	& string & reserved, optional & Value field contains a string describing the standard Open Geospatial Consortium (OGC) code \\
& & & (if any) from \citet{hare2006} corresponding to the shape and projection used in the image. \\
& & & This keyword will assure compatibility with standard GIS softwares. \\
\hline
\multicolumn{4}{l}{$^{a}$Note that IAU Working Group defines a mean radius for each Solar System body. When this Mean Radius is used in projection definition,} \\
\multicolumn{4}{l}{  the three keywords A\_RADIUS, B\_RADIUS,C\_RADIUS, must be set to the same value of the defined Mean Radius.} \\
\end{tabular}
\label{table:newkeys}
\end{sidewaystable}

@@

\item{\texttt{DATE-OBS}.
It contains the date of the observation, provided as ISO 8601
string. The value shall be assumed to refer to the start of the observation,
unless another interpretation is clearly explained in the comment field.
It is assumed to be expressed in Universal Time (UTC) but other systems are accepted as long as they
are specified in the ISO string\footnote{See table 30 in the Standard FITS document https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf}. 
PDS4 provides a \texttt{Time\_Coordinates} class with attributes
\texttt{start\_date\_time} and \texttt{stop\_date\_time} providing start and
stop date and time appropriate to the product being labeled, always in UTC.
The \texttt{time\_min} and \texttt{time\_max} fields are mandatory in the EPN-TAP
scheme, they must always be formatted in UTC as a ISO string (planned in EPN-TAP v2.1).}

@@

\item{\texttt{REFERENC}.
It contains a character string citing the bibliographic reference where the
data are first described.
It is recommended that either the Astrophysics Data System bibliographic
databases\footnote{http://adswww.harvard.edu/}
bibliographic identifier or the Digital Object Identifier\footnote{http://doi.org} (DOI)
be used.
PDS4 \texttt{External\_Reference} class and EPN-TAP \texttt{bib\_reference} have
exactly the same meaning and use than the FITS keyword.}

@@

\subsubsection{Visualization of Global and Projected Maps}
\label{sssec:maps}
High level products are meant to be used for visual inspection or GIS analysis.
They are distributed as projected rasters, often in
GeoTIFF\footnote{http://geotiff.osgeo.org/}
format, as GeoTIFF make them GIS ready.
Given some new keywords introduced in FITS and WCS definitions adapted to planetary
surface, FITS and GeoTIFF are completely interchangeable.
As an example we have downloaded a Mercury polar map from the United States Geological Survey (USGS)
PDS Annex\footnote{https://astrogeology.usgs.gov/search} and converted it in FITS.
The FITS image has been cropped in two smaller versions using the open source software
SWarp\footnote{https://www.astromatic.net/software/swarp} \citep{swarp}: SWarp was able to
automatically manage WCS planetary information.
We used ds9 to display the three FITS images and were able to correctly align them.
Linear coordinates in meters are visualized (see figure \ref{fig:ds9qgis}, left side)
as in QGIS, a popular open source GIS software, thanks to the alternate coordinate
description keywords in the header.
\begin{figure}[ht!]
\centerline{\includegraphics[height=17pc]{ess_marmo2} \includegraphics[height=17pc]{ess_marmo1}}
\caption{left: Mercury South Pole FITS images displayed in ds9, alternate coordinates in meters
are displayed;
right: GeoTIFF raster (displayed in red) and the two virtual haeders
linked to FITS rasters (displayed in green and blue); the three layers
are perfectly superposed.
Mercury image credits: MESSENGER Team, Arizona State University, Johns Hopkins Applied
Physics Laboratory, Carnegie Science}
\label{fig:ds9qgis}
\end{figure}

@@

\subsection{MATISSE}
MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool
for the Instruments for the Solar System Exploration) \citep{ZINZI2016} is the webtool
designed by the Space Science Data Center of the Italian Space Agency (SSDC-ASI) to
access, visualize and analyze data from solar system exploration missions.
Currently, MATISSE provides access to public data from five different targets (i.e., Mercury,
Venus, Mars, the dwarf planet Ceres and the asteroid Vesta).
MATISSE 2D outputs are formatted as FITS, in equirectangular projection and in planetocentric coordinates.
The planetary FITS standard described in this work has been implemented in MATISSE in order to make the output
file compliant with most FITS readers, especially DS9 and JS9.
The files generated by MATISSE could be also opened with standard GIS software (e.g.,
QGIS, ArcGIS) 
after being provided as input of the fits2vrt Python library\footnote{https://github.com/epn-vespa/fits2vrt},
whose output is a \texttt{.vrt} GDAL Virtual Header pointing to the FITS file with the appropriate
projection.
Thanks to this approach, in the next major release of MATISSE (v2.0, planned to be
online by the end of 2018) JS9 will be the standard reader for
the 2D online output of the tool, thus allowing users to perform advanced analysis in
both 2D and 3D directly inside a web-based environment.

@@

We understand that some believe FITS has grown too old for newer astronomical
applications \citep[e.g.][]{THOMAS2015133}. We are also not convinced that one format
can solve all the issues faced from long-term archiving to on-line streaming.
But we want to stress that using FITS can be an efficient compromise.
FITS can become the format that allows the astronomy and the  planetary science 
research communities to more easily share data across domains, allowing for
reduced data format conversions during processing and easy conversion for final
archival products.

@@

%
%%%%%%%%%%%%%%
% Acronyms
\begin{acronyms}
\acro{AMICA}
Asteroid Multi-band Imaging Camera
\acro{BIL}
Band Interleaved per Line
\acro{BSQ}
Band SeQuential
\acro{CRISM}
Compact Reconnaissance Imaging Spectrometer for Mars
\acro{DOI}
Digital Object Identifier
\acro{EAICD}
Experiment Archive Interface Control Document
\acro{EPN-TAP}
EuroPlaNet Table Access Protocol
\acro{ESA}
European Space Agency
\acro{FITS}
Flexible Image Transport System
\acro{GDAL}
Geospatial Data Abstraction Library
\acro{GeoTIFF}
Geographical Tagged Image File Format
\acro{GIS}
Geographical Information Systems
\acro{HiPS}
Hierarchical Progressive Survey
\acro{HiRISE}
High Resolution Imaging Science Experiment
\acro{IAU}
International Astronomical Union
\acro{IEEE}
Institute of Electrical and Electronics Engineers
\acro{IVOA}
International Virtual Observatory Alliance
\acro{ISO}
International Organization for Standardization
\acro{JAXA}
Japan Aerospace Exploration Agency
\acro{MATISSE}
Multi-purpose Advanced Tool for the Instruments for the Solar System Exploration
\acro{MEF}
Multi Extension FITS
\acro{NAVCAM}
NAVigation CAMera
\acro{NIRS}
Near-Infrared Spectrometer
\acro{OGC}
Open Geospatial Consortium
\acro{OMEGA}
Observatoire pour la Min\'eralogie, l'Eau, les Glaces et l'Activit\'e
\acro{OSIRIS}
Optical, Spectroscopic, and Infrared Remote Imaging System
\acro{PDS}
Planetary Data System
\acro{PDS3}
Planetary Data System (version 3)
\acro{PDS4}
Planetary Data System (version 4)
\acro{PSA}
Planetary Space Archive
\acro{SSDC-ASI}
Space Science Data Center of the Italian Space Agency
\acro{USGS}
United States Geological Survey
\acro{UTC}
Universal Time Coordinated
\acro{VESPA}
Virtual European Solar and Planetary Access
\acro{VO}
Virtual Observatory
\acro{WCS}
World Coordinate System
\acro{WGCCRE}
Working Group on Cartographic Coordinates and Rotational Elements
\acro{WKT}
Well Known Text
\end{acronyms}
